\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Estimating Trending Topics on Twitter with Small Subsets of the Total Data}

### 2015-01-13 paper.tex
minor fixes
updated paper more

% introduction 
Since its founding in $2006$, Twitter has grown into a corporation dominating a significant
proporition of social media today. In recent years, it has become interesting to analyze tweets
to determine interesting phenomena ranging from the viral spread of news to the detection of earthquakes \cc{Burks:2014}. The volume of existing tweets is prohibitively large for standard methods of analysis, and requires approaches that are able to either store large amounts of data in memory for fast access (including parallel approaches to data processing) or approaches that only require sublinear memory while retaining accuracy guarantees. In this paper, we investigate a problem utilizing the latter method.

@@

More specifically, we will attempt to approximate the $k$ most popular Twitter hashtags in the time intervals of length $y$ in order to tell what is trending during that time period.
The idea is that estimating the top $k$ hashtags in some time interval provides a model for the topics that are trending on Twitter in that time period.

@@

Analyzing what is trending on Twitter is a more recent research area. Twitter itself must give its users hashtags which it deems trending, to both spread the hashtag and to provide an awareness of what is going on in the world to its users. Knowledge of what is trending and how that distribution looks over time is also useful for the business model, which relies on ads to a certain extent.
Twitter itself stores all of its frequency data, and looks at recent spikes while conditioning on all of the past frequency data in order to determine estimates of trending likelihood.
In $2012$, MIT Professor Devavrat Shah built a machine-learning algorithm to predict which topics will trend on Twitter an hour and a half before they do, with $95\%$ accuracy, though the datasets they tested their algorithm on were small\cc{MITNews2012}.

@@

The reason why the history is important is as follows: Some hashtags are constantly present. For example, in our experiments, the hashtags $\#$porn, $\#$gameinsight, $\#$teamfollowback had nearly always had a high frequency in the current time unit. This fact is not surprising given the nature of the first hashtag. The second hashtag, $\#gameinsight$, is the name of popular video game maker that people tweet when they win achievements in the game. Finally, $\#teamfollowback$ is a method Twitter users apply to gather Twitter followers quickly. The rule is if you follow that tag, you follow people who post that tag. Since these sorts of tags are consistently present, they should not be consider trending: they are not new. Therefore, we need to filter these out with the history.
The necessity of maintaining approximate counts for a longer history depends on exactly how constant these consistently-appearing tweets actually are.

@@

To store data about all occurrences of every hashtag seen in the past, we use a modified version of the time-aggregated Hokusai system \cc{Matusevych:2012}, which is an extension of the Count-Min sketch. We previously described this data structure in \nref{sec:PreviousWork}. 
To the Hokusai structure we add another Count-Min sketch that combines the information from the Hokusai Count-Min sketches. We call this external Count-Min sketch the Kernel, since it acts as a weighting function on the CM sketches in the Hokusai structure. Its role is to depreciate the value of older data.
Denoting the CM sketches of the Hokusai structure as a vector $\textbf{M} = \{\overline{M}, M^0, M^1, ..., M^{{\lfloor {\log(T)} \rfloor}}\}$, where $\overline{M}$ is the Count-Min sketch storing the data of a unit time step (of size $z$), $M^j$ is the Count-Min sketch storing the data for the sketch with a $2^j$ time resolution, and $T$ is an upper bound on the number of $z$-units stored by the data structure. Then, the kernel function is given by

@@

\begin{algorithm} 
\caption{Update History} \label{alg1}
\begin{algorithmic}[1]
\ForAll {$i$}
	\State Initialize Count-Min sketch $M^i = 0$
\EndFor
\State Initialize $t = 0$
\State Initialize Count-Min sketches $\overline{M} = 0$ and $K = 0$
\While {data arrives}
	\State Aggregate data into sketch $\overline{M}$ for current $z$-unit while also adding this data to $K$
	\State $t \leftarrow t + 1$ (increment counter)
	\State $K \leftarrow K - \overline{M}$
	\For {$j = 0$ to argmax \{$l$ where $t$ mod $2^l = 0$\}}
		\State $K \leftarrow K + 2^{-j}(\overline{M} - M^j)$
		\State $T \leftarrow \overline{M}$ (back up temporary storage)
		\State $\overline{M} \leftarrow \overline{M} + M^j$ (increment cumulative sum)
		\State $M^j \leftarrow T$ (new value for $M^j$)
	\EndFor
	\State $\overline{M} \leftarrow 0$ (reset aggregator)
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm} 
\caption{Update Current-Window} \label{alg2}
\begin{algorithmic}[1]
\While {data arrives}
	\If {the current $z$-unit is different than that of the last hashtag seen}
		\State Do all the end-of-$z$ aggregation for the History structure as detailed in 
		Algorithm \aref{alg1}.
		\ForAll {elements in the Fibonacci heap}
			\State look up the hashtag corresponding to this node in the hash table
			\State Update the key of the node to: \[\frac{\textnormal{frequency in last } y-\textnormal{length time period found at the table entry}}{\textnormal{new value in History data structure}}\]
		\EndFor
	\EndIf
	\If {queue is not empty}
		\State Peek at end of queue.
		\If {the timestamp + $y$ is before the current time}
			\State Look up the hashtag in the hash table and decrement the stored frequency.
			\If {the frequency is now 0}
				\State Delete the node in the heap pointed to by this entry in the table.
				\State Delete this entry in the hash table.
			\Else	
				\State Update the key of this node pointed to by this entry in the table to the proper value given the new frequency.
			\EndIf
		\EndIf
	\EndIf
	\If {hashtag is in hash table}
		\State Increment the frequency stored at that entry.
		\State Update the key of the node in the Fibonacci heap.
	\Else
		\State Insert a new node into the Fibonacci heap with the appropriate key and value. 
		\State Insert the hashtag into the hash table with a pointer to this node.
 	\EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

We plot this data in the histogram shown in \aref{fig:jaccard_sims}. The $x$-axis is the Jaccard similarity, and the $y$-axis is the count. The mean was $0.47$, and the standard Deviation was $0.18$. From the plot, we can see that the distribution of the Jaccard similarities is roughly a centered normal distribution around $0.5$. A Jaccard similarity of $0.5$ means about half the hashtags are the same in each $3$-hour time interval between the two algorithms.

@@

Therefore, given some adjustments, the algorithms developed in this paper open the doors to new exploration of topic modeling the Twitter data stream.

@@

With respect to the algorithms specifically, in the future we would want to adapt our approach to the history-sensitive algorithm and play with the parameters a bit more. We would also like to try different kernel functions to see how they performed. Another approach to utilizing the history information would be instead of weighting by inverse exponentially weighted frequency in the past for a given hashtag, attempt to calculate some score of periodicity for the hashtag. If the hashtag has low periodicity over a given time scale, then even if it appears multiple times in the past (because the past is long), our algorithm will avoid discarding it as non-trending.

@@

We provide a link to \href{https://github.com/kiranvodrahalli/cos521/}{all our code}, as well as to an online hosting of the \href{http://www.princeton.edu/~awlee/trends.html}{frequency visualization}. In the repository, there are also the results files, Powerpoint slides, and citations for the pictures we used.

@@

\section{Appendix II: Top-$k$ Hashtags in $6$ Interesting Time Chunks} \label{sec:Appendix_topk}

>>>>

% introduction 
Since its founding in $2006$, Twitter has grown into a corporation dominating a significant
proportion of social media today. In recent years, it has become interesting to analyze tweets
to determine interesting phenomena ranging from the viral spread of news to the detection of earthquakes \cc{Burks:2014}. The volume of existing tweets is prohibitively large for standard methods of analysis, and requires approaches that are able to either store large amounts of data in memory for fast access (including parallel approaches to data processing) or approaches that only require sublinear memory while retaining accuracy guarantees. In this paper, we investigate a problem utilizing the latter method.

@@

More specifically, we will attempt to approximate the $k$ most popular Twitter hashtags in time intervals of length $y$ in order to tell what is trending during that time period.
The idea is that estimating the top $k$ hashtags in some time interval provides a model for the topics that are trending on Twitter in that time period.

@@

Analyzing what is trending on Twitter is a more recent research area. Twitter itself must give its users hashtags which it deems trending, to both spread the hashtag and to provide an awareness of what is going on in the world to its users. Knowledge of what is trending and how that distribution looks over time is also useful for the business model, which relies on ads to a certain extent.
Twitter itself stores all of its frequency data, and looks at recent spikes while conditioning on all of the past frequency data in order to determine estimates of trending likelihood.
In $2012$, MIT Professor Devavrat Shah built a machine-learning algorithm to predict which topics will trend on Twitter an hour and a half before they do, with $95\%$ accuracy, though the datasets they tested their algorithm on were small \cc{MITNews2012}.

@@

The reason why the history is important is as follows: Some hashtags are constantly present. For example, in our experiments, the hashtags $\#porn$, $\#gameinsight$, $\#teamfollowback$ had nearly always had a high frequency in the current time unit. This fact is not surprising given the nature of the first hashtag. The second hashtag, $\#gameinsight$, is the name of popular video game maker that people tweet when they win achievements in the game. Finally, $\#teamfollowback$ is a method Twitter users apply to gather Twitter followers quickly. The rule is if you follow that tag, you follow people who post that tag. Since these sorts of tags are consistently present, they should not be considered trending: they are not new. Therefore, we need to filter these out with the history.
The necessity of maintaining approximate counts for a longer history depends on exactly how constant these consistently-appearing tweets actually are.

@@

To store data about all occurrences of every hashtag seen in the past, we use a modified version of the time-aggregated Hokusai system \cc{Matusevych:2012}, which is an extension of the Count-Min sketch. We already described this data structure in \nref{sec:PreviousWork}. 
To the Hokusai structure we add another Count-Min sketch that combines the information from the Hokusai Count-Min sketches. We call this external Count-Min sketch the Kernel, since it acts as a weighting function on the CM sketches in the Hokusai structure. Its role is to depreciate the value of older data.
We denote the CM sketches of the Hokusai structure as a vector $\textbf{M} = \{\overline{M}, M^0, M^1, ..., M^{{\lfloor {\log(T)} \rfloor}}\}$, where $\overline{M}$ is the Count-Min sketch storing the data of a unit time step (of size $z$), $M^j$ is the Count-Min sketch storing the data for the sketch with a $2^j$ time resolution, and $T$ is an upper bound on the number of $z$-units stored by the data structure. Then, the kernel function is given by

@@

\begin{algorithm} 
\caption{Update History} \label{alg1}
\begin{algorithmic}[1]
\ForAll {$i$}
	\State Initialize Count-Min sketch $M^i = 0$
\EndFor
\State Initialize $t = 0$
\State Initialize Count-Min sketches $\overline{M} = 0$ and $K = 0$
\While {data arrives}
	\State Aggregate data into sketch $\overline{M}$ for current $z$-unit
	\State $t \leftarrow t + 1$ (increment counter)
	\For {$j = 0$ to argmax \{$l$ where $t$ mod $2^l = 0$\}}
		\State $K \leftarrow K + 2^{-j}(\overline{M} - M^j)$ (update the Kernel)
		\State $T \leftarrow \overline{M}$ (back up temporary storage)
		\State $\overline{M} \leftarrow \overline{M} + M^j$ (increment cumulative sum)
		\State $M^j \leftarrow T$ (new value for $M^j$)
	\EndFor
	\State $\overline{M} \leftarrow 0$ (reset aggregator)
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm} 
\caption{Update Current-Window} \label{alg2}
\begin{algorithmic}[1]
\While {data arrives}
	\If {the current $z$-unit is different than that of the last hashtag seen}
		\State Do all the end-of-$z$ aggregation for the History structure as detailed in 
		Algorithm \aref{alg1}.
		\ForAll {elements in the Fibonacci heap}
			\State look up the hashtag corresponding to this node in the hash table
			\State Update the key of the node to: $\textbf{priority}(\textnormal{hashtag})$
		\EndFor
	\EndIf
	\If {queue is not empty}
		\State Peek at end of queue.
		\If {the timestamp + $y$ is before the current time}
			\State Look up the hashtag in the hash table and decrement the stored frequency.
			\If {the frequency is now 0}
				\State Delete the node in the heap pointed to by this entry in the table.
				\State Delete this entry in the hash table.
			\Else	
				\State Update the key of the node with the new frequency.
			\EndIf
		\EndIf
	\EndIf
	\If {hashtag is in hash table}
		\State Increment the frequency stored at that entry.
		\State Update the key of the node in the Fibonacci heap.
	\Else
		\State Insert a new node into the Fibonacci heap with the appropriate key and value. 
		\State Insert the hashtag into the hash table with a pointer to this node.
 	\EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

We plot this data in the histogram shown in \aref{fig:jaccard_sims}. The $x$-axis is the Jaccard similarity, and the $y$-axis is the count. The mean was $0.47$, and the standard deviation was $0.18$. From the plot, we can see that the distribution of the Jaccard similarities is roughly a centered normal distribution around $0.5$. A Jaccard similarity of $0.5$ means about half the hashtags are the same in each $3$-hour time interval between the two algorithms.

@@

Therefore, given some adjustments, the algorithms developed in this paper open the doors to new explorations of topic modeling the Twitter data stream.

@@

With respect to the algorithms specifically, in the future we would want to adapt our approach to the history-sensitive algorithm and play with the parameters a bit more. We would also like to try different kernel functions to see how they performed. Another approach to utilizing the history information would be instead of weighting by inverse-exponentially-weighted frequency in the past for a given hashtag, attempt to calculate some score of periodicity for the hashtag. If the hashtag has low periodicity over a given time scale, then even if it appears multiple times in the past, our algorithm will avoid discarding it as non-trending despite the potentially high frequency count (because the past is long).

@@

We provide a link to \href{https://github.com/kiranvodrahalli/cos521/}{all our code}. In the repository, there are also the results files, Powerpoint slides, and citations for the pictures we used.

We also provide a link to an online hosting of the \href{http://www.princeton.edu/~awlee/trends.html}{frequency visualization}. The frequency visualization is a histogram of hashtag frequencies, displayed over time. We cherry-picked a set of interesting hashtags based on our observations, and then made this visualization to present the changes in frequency in $3$-hour intervals over time of these hashtags.

The fact that these frequencies change over time and spike and fall in turns suggests that
when the hashtag frequencies spike, the hashtags are trending. The goal of our algorithms is to identify the hashtags (of a much larger set) which are trending at a given point in time.

So we hope that for a selected $3$-hour interval, the hashtags that our algorithms say are trending include the ones that are trending at that time in this visualization.

@@

\section{Appendix II: Top-$k$ Hashtags Results} \label{sec:Appendix_topk}
